An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-06-21 cal_demo/ms.tex
More fine tuning as I re-read referee reports. Also updating .pdf

Traditional radio correlators cross-multiply the signals from all pairs of antennas, and 
the computation scales as the number of antennas squared, $\mathcal{O}(\Nant^2)$ 
\citep{bun04}. As the number of elements in future arrays grows, the computational cost will 
become prohibitively expensive, and exploring efficient correlator schemes is essential to 
enable next generation instruments \citep{lon00}. Meanwhile, radio transient monitoring 
requires access to high time and frequency resolution data to identify and characterize events 
such as fast radio bursts (FRBs, \citealt{lor07}), or to follow up gravitational wave candidates 
with radio observations \citep{abb16a,abb16b}. FRBs are highly unexplored at low frequencies 
(< 1 GHz), but are expected to occur on timescales $\Delta t \sim$ 1--10~ms \citep{tho13}. 
Recording the full visibility matrix for $\Nant \gtrsim 10^3$ arrays at this timescale leads to 
extremely high data write rates.

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional cross correlator, pair-wise visibilities are written to disk and used to calibrate 
offline before further processing such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

Thus far we have assumed the antennas perfectly measure the sky electric field distribution according to their far-field radiation patterns. Here we introduce a multiplicative complex gain as well as an additive noise term which corrupt the measured signals.
\begin{equation}\label{eq:apply_gain}
E_a(f,t) = g_a(f,t) E_a^I(f,t) + n_a(f,t)
\end{equation}
%The noise term is 
%strictly receiver noise -- noise introduced by the instrument. Any sky noise is implicitly included 
%in the time dependence of the sky electric field. 
We have carried the frequency dependence to this point to make clear the complex gains are
frequency dependent. However, each subsequent step treats each frequency 
channel independently, so we will drop the $f$ to simplify notation. 
Furthermore, while antenna gains may vary with time, we will assume they are constant over
the period of our calibration, and thus drop their time dependence from the notation.

@@

The pipeline up to this point can be summarized 
through the following equation.
\begin{align}
\mathcal{E}'(\s_k,t) & = \underbrace{\frac{1}{\Nant} \sum_j e^{2\pi i \mathbf{r}_j \cdot \s_k}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a W_a(\mathbf{r}_j - \ra)}_{\mathrm{Gridding}} \\
& \qquad \times \underbrace{ \frac{1}{g'_a} \left(g_a E_a^I(t)+n_a(t)\right)}_{\mathrm{Calibration}} \nonumber
\end{align}
The sum over antennas uses the antenna illumination pattern to grid the calibrated electric field measurements onto 
regular gridpoints, $\mathbf{r}_j$. The sum over $j$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums and use the DFT to approximate the Fourier relationship between
the antenna illumination pattern and the far-field radiation pattern.
\begin{align}\label{eq:epix}
\mathcal{E}(\s_k,t) & = \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_a E^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra} \nonumber \\
  & \qquad\times \sum_j W_a(\mathbf{r}_j-\ra)e^{2\pi i \s_k \cdot (\mathbf{r}_j-\ra)} \nonumber\\
& \approx \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_aE^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra}\mathcal{W}_a(\s_k)
\end{align}
In this form we see that the effect of gridding with the antenna illumination pattern is to attenuate 
the image by a factor of the far-field radiation response, $\mathcal{W}_a(\s)$.

Finally, the images are squared and averaged in time to reduce output bandwidth, then written 
to disk. While operations have been reordered, the resulting image is equivalent to one 
produced by gridding and imaging visibilities. By shifting from cross-correlation of all pairs of 
antennas to a spatial FFT, the computational cost is significantly lowered for certain classes of 
compact arrays. The output bandwidth is also greatly reduced as visibilities are exchanged for 
gridded images. A full exploration of these scalings was presented in \citealt{thy15c}.

>>>>

The most common radio correlator designs of today, the FX and lag correlators \citep{rom99},
cross-multiply the signals from all pairs of antennas, and 
the computation scales as the number of antennas squared, $\mathcal{O}(\Nant^2)$ 
\citep{bun04}. As the number of elements in future arrays grows, the computational cost will 
become prohibitively expensive, and exploring efficient correlator schemes is essential to 
enable next generation instruments \citep{lon00}. Meanwhile, radio transient monitoring 
requires access to high time and frequency resolution data to identify and characterize events 
such as fast radio bursts (FRBs, \citealt{lor07}), or to follow up gravitational wave candidates 
with radio observations \citep{abb16a,abb16b}. FRBs are highly unexplored at low frequencies 
(< 1 GHz), but are expected to occur on timescales $\Delta t \sim$ 1--10~ms \citep{tho13}. 
Recording the full visibility matrix for $\Nant \gtrsim 10^3$ arrays at this timescale leads to 
extremely high data write rates.

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional cross correlator, pair-wise visibilities are written to disk and used to calibrate 
after observing before further processing such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

Thus far we have considered the perfectly measured electric field distribution according to 
the antenna far-field radiation patterns. Here we introduce a multiplicative complex gain as 
well as an additive noise term which corrupt the measured signals.
\begin{equation}\label{eq:apply_gain}
E_a(f,t) = g_a(f,t) E_a^I(f,t) + n_a(f,t)
\end{equation}
%The noise term is 
%strictly receiver noise -- noise introduced by the instrument. Any sky noise is implicitly included 
%in the time dependence of the sky electric field. 
We have carried the frequency dependence to this point to make clear the complex gains are
frequency dependent. However, each subsequent step treats each frequency 
channel independently, so we will drop the $f$ to simplify notation. 
Furthermore, while antenna gains may vary with time, we will assume they are constant over
the period of our calibration, and thus drop their time dependence from the notation.

@@

The pipeline up to this point can be summarized 
through the following equation.
\begin{align}
\mathcal{E}'(\s_k,t) & = \underbrace{\frac{1}{\Nant} \sum_j e^{2\pi i \mathbf{r}_j \cdot \s_k}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a W_a(\mathbf{r}_j - \ra)}_{\mathrm{Gridding}} \\
& \qquad \times \underbrace{ \frac{1}{g'_a} \left(g_a E_a^I(t)+n_a(t)\right)}_{\mathrm{Calibration}} \nonumber
\end{align}
The sum over antennas uses the antenna illumination pattern to grid the calibrated electric field measurements onto 
regular gridpoints, $\mathbf{r}_j$. The sum over $j$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums and use the DFT to approximate the Fourier relationship between
the antenna illumination pattern and the far-field radiation pattern.
\begin{align}\label{eq:epix}
\mathcal{E}'(\s_k,t) & = \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_a E^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra} \nonumber \\
  & \qquad\times \sum_j W_a(\mathbf{r}_j-\ra)e^{2\pi i \s_k \cdot (\mathbf{r}_j-\ra)} \nonumber\\
& \approx \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_aE^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra}\mathcal{W}_a(\s_k)
\end{align}
In this form we see that the effect of gridding with the antenna illumination pattern is to attenuate 
the image by a factor of the far-field radiation response, $\mathcal{W}_a(\s)$.

Finally, the images are squared and averaged in time to form dirty images, 
$I'(\hat{\mathbf{s}}_k)$, then written 
to disk. While operations have been reordered, the resulting image is equivalent to one 
produced by gridding and imaging visibilities. By shifting from cross-correlation of all pairs of 
antennas to a spatial FFT, the computational cost is significantly lowered for certain classes of 
compact arrays. The output bandwidth is also greatly reduced as visibilities are exchanged for 
gridded images. A full exploration of these scalings was presented in \citealt{thy15c}.
